---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Loading and preprocessing the data
```{r}
# Setting working directory
# setwd("~/Projetos em R/Coursera Reproducible Research/Week 2")
data <- read.csv("activity.csv", header=TRUE, na.strings="NA")
data <- transform(data, date = as.Date(date))
head(data)
summary(data)
```

## 2. Getting the mean total number of steps taken per day
### 2.1. Calculating the total number of steps taken per day
```{r}
steps <- aggregate(data$steps, by=list(Date=data$date), FUN=sum)
names(steps)[names(steps) == "x"] <- "Total"
head(steps)
```

### 2.2. Making a histogram of the total number of steps taken each day
```{r}
library(ggplot2)
hist <- ggplot(data=na.omit(steps), aes(Total)) +
  geom_histogram(binwidth=1500, colour="red", fill="white") + 
  xlab("Total Number of Steps Taken Each Day") +
  ylab("Count") +
  ggtitle("Histogram of the Total Number of Steps Taken Each Day")
print(hist)
```

### 2.3. Calculating and reporting the mean and median of the total number of steps taken per day
```{r}
mean(na.omit(steps$Total))
median(na.omit(steps$Total))
```

## 3. Getting the average daily activity pattern
### 3.1. Making a time series plot of the 5-minute interval (x-axis) and the average number os steps taken, averaged across all days (y-axis)
```{r}
steps_by_min <- aggregate(steps ~ interval, data=data, FUN=mean)
time_series_plot <- ggplot(data=steps_by_min, aes(x=interval, y=steps)) + 
        geom_line() +
        xlab("Time Intervals (5-Minute)") +
        ylab("Total Number of Steps") +
        ggtitle("Average Number of Steps Taken of the 5-Minute Interval")
print(time_series_plot)
```

### 3.2. Getting the 5-minute interval that, on average across all the days in the dataset, contains the maximum number of steps
```{r}
steps_by_min[which(steps_by_min$steps == max(steps_by_min$steps)),]
```
## 4. Imputing missing values
### 4.1. Calculating and reporting the total number of missing values in the dataset
```{r}
sapply(X=data, FUN=function(x) sum(is.na(x)))
```

### 4.2. Devising a strategy for filling in all of the missing values in the dataset
#### Strategy: Use the mean for 5-minute interval to replace all the missing values in the dataset
```{r}
library(dplyr)
data_NA_replace <- function(x) replace(x, is.na(x), mean(x, na.rm=TRUE))
```

### 4.3. Creating a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
data_without_NA <- data%>% group_by(interval) %>% mutate(steps=data_NA_replace(steps))
data_new <- as.data.frame(data_without_NA)
head(data_new)
summary(data_new)
```

#### Counting missing values, now
```{r}
sum(is.na(data_new))
```

### 4.4. Making a histogram of the total number of steps taken each day, calculating, and reporting the mean and median total number of steps taken per day
```{r}
steps_new <- aggregate(data_new$steps, by=list(data_new$date), FUN=sum)
names(steps_new)[names(steps_new) == "x"] <- "Total"
names(steps_new)[names(steps_new) == "Group.1"] <- "Date"
hist_new <- ggplot(data=steps_new, aes(Total)) +
          geom_histogram(binwidth=1500, colour="red", fill="white") + 
  xlab("Total Number of Steps Taken Each Day") +
  ylab("Count") +
  ggtitle("Histogram of the Total Number of Steps Taken Each Day with no missing values")
print(hist_new)
```

#### Comparison and discussion about the data:
```{r}
mean(na.omit(steps$Total))
median(na.omit(steps$Total))
mean(steps_new$Total)
median(steps_new$Total)
```
#### The means of each dataset are same. The medians of each dataset are different.Therefore, the impact of imputing missing data on the estimates of the total daily number of steps has decayed on the value that separating the higher half from the lower half of dataset.

## 5. Getting the differences in activity patterns between weekdays and weekends
### 5.1. Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day
```{r}
data_new$date <- as.Date(data_new$date)
data_new$WeekdayOrWeekend <- ifelse(weekdays(data_new$date) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend")
head(data_new)
```

### 5.2. Making a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number os steps taken, averaged across all days or weekend days (y-axis)
```{r}
data_new <- (data_new %>% group_by(interval, WeekdayOrWeekend) %>% summarise(Mean=mean(steps)))
ggplot(data_new, mapping=aes(x=interval, y=Mean)) +
        geom_line() +
        facet_grid(WeekdayOrWeekend ~.) +
        xlab("Interval") +
        ylab("Mean of Steps") +
        ggtitle("Comparison of Average Number of Steps in Each Inverval")
```

#### The plots shows differences in activity patterns between weekdays and weekends. During weekdays, people tend to be more active; during weekend, people tend to get up late and stay up late.